Full Dissertation

### 2024-04-06 Sections/full.qmd
turn off external-filter
render full in pdf and markdown

---
title: Full Dissertation
subtitle: An instance-based model account of the benefits of varied practice in visuomotor skill
date: 04/04/24
#page-layout: full
code-fold: true
code-tools: true
toc: true
toc-location: body
toc-title: Contents
execute: 
  warning: false
  eval: true
format:
  html: default
  hugo-md:
    echo: false
    output: false
    html-math-method: mathjax
    output-file: diss_full_hugo.md
  gfm:
    echo: true
    output-file: diss_full_gfm.md
  hikmah-manuscript-pdf:
    echo: false
    output-file: "tg_full_disser.pdf"
    mainfont: "Linux Libertine O"
    mainfontoptions:
      - "Numbers=Proportional"
      - "Numbers=OldStyle"
    mathfont: "Libertinus Math"
editor: 
  markdown: 
    wrap: 72
prefer-html: true
---

@@

@@

Research on the effects of varied training typically compares participants trained under constant, or minimal variability conditions to those trained from a variety of examples or conditions [@czyzVariabilityPracticeInformation2021; @soderstromLearningPerformanceIntegrative2015]. Varied training has been shown to influence learning in myriad domains including categorization of simple stimuli [@hahnEffectsCategoryDiversity2005; @maddoxStimulusRangeDiscontinuity2011; @posnerGenesisAbstractIdeas1968], complex categorization [@nosofskyModelguidedSearchOptimal2018], language learning [@jonesDensityDistinctivenessEarly2020; @perryLearnLocallyThink2010; @twomeyAllRightNoises2018; @wonnacottInputEffectsAcquisition2012], anagram completion [@goodeSuperiorityVariableRepeated2008], trajectory extrapolation [@fulvioTaskSpecificResponseStrategy2014], task switching [@sabahWhenLessMore2019], associative learning [@leeEvidentialDiversityIncreases2019], visual search [@georgeStimulusVariabilityTask2021; @gonzalezDiversityTrainingEnhances2011; @kelleyLearningAttendEffects2009], voice identity learning [@lavanEffectsHighVariability2019], simple motor learning [@braunMotorTaskVariation2009; @kerrSpecificVariedPractice1978; @rollerVariablePracticeLenses2001; @willeyLimitedGeneralizationVaried2018], sports training  [@greenPracticeVariabilityTransfer1995a; @northEffectConsistentVaried2019], and training on a complex video game [@seowTransferEffectsVaried2019].

Training variation has received a particularly large amount of attention within the domain of visuomotor skill learning. Much of this research has been influenced by the work of @schmidtSchemaTheoryDiscrete1975, who proposed a schema-based account of motor learning as an attempt to address the longstanding problem of how novel movements are produced. According to Schema Theory, learners possess general motor programs for classes of movements (e.g. throwing a ball with an underhand movement), as well as schema rules that determine how a motor program is parameterized or scaled for a particular movement. Schema theory predicts that varied training results in the formation of a more general schema-rule, which can allow for transfer to novel movements within a given movement class. Experiments that test this hypothesis are often designed to compare the transfer performance of a constant-trained group against that of a varied-trained group. Both groups train on the same task, but the varied group practices from multiple levels of a task-relevant dimension that remains invariant for the constant group. For example, investigators might train two groups of participants to throw a projectile at a target, with a constant group that throws from a single location, and a varied group that throws from multiple locations. Both groups are then tested from novel locations. Empirically observed benefits of the varied-trained group are then attributed to the variation they received during training, a finding observed in numerous studies [@catalanoDistantTransferCoincident1984a; @chuaPracticeVariabilityPromotes2019; @goodwinEffectDifferentQuantities1998; @kerrSpecificVariedPractice1978; @wulfEffectTypePractice1991], and the benefits of this variation are typically thought to be mediated by the development of a more general schema for the throwing motion.

@@

To prepare the data, we first removed trials that were not easily
interpretable as performance indicators in our task. Removed trials
included: 1) those in which participants dragged the ball outside of the
orange starting box without releasing it, 2) trials in which
participants clicked on the ball, and then immediately released it,
causing the ball to drop straight down, 3) outlier trials in which the
ball was thrown more than 2.5 standard deviations further than the
average throw (calculated separately for each throwing position), and 4)
trials in which the ball struck the barrier. The primary measure of
performance used in all analyses was the absolute distance away from the
center of the target. The absolute distance was calculated on every
trial, and then averaged within each subject to yield a single
performance score, for each position. A consistent pattern across
training and testing phases in both experiments was for participants to
perform worse from throwing positions further away from the target -- a
pattern which we refer to as the difficulty of the positions. However,
there were no interactions between throwing position and training
conditions, allowing us to collapse across positions in cases where
contrasts for specific positions were not of interest. All data
processing and statistical analyses were performed in R version 4.03 (R
Core Team, 2020). ANOVAs for group comparisons were performed using the
rstatix package (Kassambara, 2021)****.

@@

It is common for psychological process models of categorization learning
to use an approach such as multidimensional scaling so as to transform
the stimuli from the physical dimensions used in the particular task
into the psychological dimensions more reflective of the actual human
representations [@nosofskySimilarityScalingCognitive1992; @shepardUniversalLawGeneralization1987]. Such scaling typically entails having participants rate the similarity between individual items
and using these similarity judgements to then compute the psychological
distances between stimuli, which can then be fed into a subsequent
model. In the present investigation, there was no such way to scale the
x and y velocity components in terms of the psychological similarity,
and thus our modelling does rely on the assumption that the
psychological distances between the different throwing positions are
proportional to absolute distances in the metric space of the task (e.g.
the relative distance between positions 400 and 500 is equivalent to
that between 800 and 900). However, an advantage of our approach is that
we are measuring similarity in terms of how participants behave
(applying a velocity to the ball), rather than the metric features of
the task stimuli.

## Conclusion

@@

```{dot}
//| label: fig-design-e1
//| fig-cap: "Experiment 1 Design. Constant and Varied participants complete different training conditions."
//| fig-width: 7.0
//| fig-height: 2.5
//| fig-responsive: false
/////| column: screen-inset-right
digraph {
  graph [layout = dot, rankdir = LR]

@@

  Test2 -> Test3
}
```

@@

```{dot}
//| label: fig-design-e2
//| fig-cap: "Experiment 2 Design. Constant and Varied participants complete different training conditions. The training and testing bands are the reverse of Experiment 1. "
//| fig-width: 8.0
//| fig-height: 2.5
//| fig-responsive: false
//| 
//| 
/////| column: screen-inset-right
digraph {
  graph [layout = dot, rankdir = LR]
  node [shape = rectangle, style = filled]
  data1 [label = " Varied Training \n100-300\n350-550\n600-800", fillcolor = "#FF0000"]
  data2 [label = " Constant Training \n600-800", fillcolor = "#00A08A"]
  Test3 [label = "    Final Test \n  Novel With Feedback  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
  data1 -> Test1
  data2 -> Test1
  subgraph cluster {
    label = "Test Phase \n(Counterbalanced Order)"
    Test1 [label = "Test  \nNovel Bands  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
    Test2 [label = "  Test \n  Varied Training Bands  \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]
    Test1 -> Test2
  }
  Test2 -> Test3
}
```

@@

\ \ \ \

>>>>

---
title: Full Dissertation
subtitle: "The Role of Variability in Learning Transfer: A Similarity-Based Computational Approach"
date: 04/04/24
author:
- name: Thomas Gorman
  affiliation: Indiana University
page-layout: full
code-fold: true
code-tools: true
toc: true
toc-location: body
toc-title: Contents
cache: true
execute: 
  warning: false
  eval: true
format:
  html: default
  hugo-md:
    echo: false
    include: false
    html-math-method: mathjax
    output-file: diss_full_hugo.md
  gfm:
    echo: false
    output: false
    output-file: diss_full_gfm.md
  hikmah-manuscript-pdf:
    echo: false
    output-file: "tg_full_disser.pdf"
    mainfont: "Linux Libertine O"
    mainfontoptions:
      - "Numbers=Proportional"
      - "Numbers=OldStyle"
    mathfont: "Libertinus Math"
editor: 
  markdown: 
    wrap: 72
prefer-html: true
---

@@

```{r}
#| cache: false

@@

Research on the effects of varied training typically compares participants trained under constant, or minimal variability conditions to those trained from a variety of examples or conditions [@czyzVariabilityPracticeInformation2021; @soderstromLearningPerformanceIntegrative2015]. Varied training has been shown to influence learning in myriad domains including categorization of simple stimuli [@hahnEffectsCategoryDiversity2005; @maddoxStimulusRangeDiscontinuity2011; @posnerGenesisAbstractIdeas1968], complex categorization [@nosofskyModelguidedSearchOptimal2018], language learning [@jonesDensityDistinctivenessEarly2020; @perryLearnLocallyThink2010; @twomeyAllRightNoises2018; @wonnacottInputEffectsAcquisition2012], anagram completion [@goodeSuperiorityVariableRepeated2008], trajectory extrapolation [@fulvioTaskSpecificResponseStrategy2014], task switching [@sabahWhenLessMore2019], associative learning [@leeEvidentialDiversityIncreases2019], visual search [@georgeStimulusVariabilityTask2021; @gonzalezDiversityTrainingEnhances2011; @kelleyLearningAttendEffects2009], voice identity learning [@lavanEffectsHighVariability2019], simple motor learning [@braunMotorTaskVariation2009; @kerrSpecificVariedPractice1978; @rollerVariablePracticeLenses2001; @willeyLimitedGeneralizationVaried2018], sports training  [@greenPracticeVariabilityTransfer1995a; @northEffectConsistentVaried2019], and training on a complex video game [@seowTransferEffectsVaried2019].
Training variation has received a particularly large amount of attention within the domain of visuomotor skill learning. Much of this research has been influenced by the work of @schmidtSchemaTheoryDiscrete1975, who proposed a schema-based account of motor learning as an attempt to address the longstanding problem of how novel movements are produced. According to Schema Theory, learners possess general motor programs for classes of movements (e.g. throwing a ball with an underhand movement), as well as schema rules that determine how a motor program is parameterized or scaled for a particular movement. Schema theory predicts that varied training results in the formation of a more general schema-rule, which can allow for transfer to novel movements within a given movement class. Experiments that test this hypothesis are often designed to compare the transfer performance of a constant-trained group against that of a varied-trained group. Both groups train on the same task, but the varied group practices from multiple levels of a task-relevant dimension that remains invariant for the constant group. For example, investigators might train two groups of participants to throw a projectile at a target, with a constant group that throws from a single location, and a varied group that throws from multiple locations. Both groups are then tested from novel locations. Empirically observed benefits of the varied-trained group are then attributed to the variation they received during training, a finding observed in numerous studies [@catalanoDistantTransferCoincident1984a; @chuaPracticeVariabilityPromotes2019; @goodwinEffectDifferentQuantities1998; @kerrSpecificVariedPractice1978; @wulfEffectTypePractice1991], and the benefits of this variation are typically thought to be mediated by the development of a more general schema for the throwing motion.

@@

To prepare the data, we first removed trials that were not easily interpretable as performance indicators in our task. Removed trials included: 1) those in which participants dragged the ball outside of the orange starting box without releasing it, 2) trials in which participants clicked on the ball, and then immediately released it, causing the ball to drop straight down, 3) outlier trials in which the ball was thrown more than 2.5 standard deviations further than the average throw (calculated separately for each throwing position), and 4) trials in which the ball struck the barrier. The primary measure of performance used in all analyses was the absolute distance away from the center of the target. The absolute distance was calculated on every trial, and then averaged within each subject to yield a single performance score, for each position. A consistent pattern across training and testing phases in both experiments was for participants to perform worse from throwing positions further away from the target -- a pattern which we refer to as the difficulty of the positions. However, there were no interactions between throwing position and training conditions, allowing us to collapse across positions in cases where contrasts for specific positions were not of interest. All data processing and statistical analyses were performed in R version 4.03 (R Core Team, 2020). ANOVAs for group comparisons were performed using the rstatix package (Kassambara, 2021).

@@

It is common for psychological process models of categorization learning to use an approach such as multidimensional scaling so as to transform the stimuli from the physical dimensions used in the particular task into the psychological dimensions more reflective of the actual human representations [@nosofskySimilarityScalingCognitive1992; @shepardUniversalLawGeneralization1987]. Such scaling typically entails having participants rate the similarity between individual items and using these similarity judgements to then compute the psychological distances between stimuli, which can then be fed into a subsequent model. In the present investigation, there was no such way to scale the x and y velocity components in terms of the psychological similarity, and thus our modelling does rely on the assumption that the psychological distances between the different throwing positions are proportional to absolute distances in the metric space of the task (e.g. the relative distance between positions 400 and 500 is equivalent to that between 800 and 900). However, an advantage of our approach is that we are measuring similarity in terms of how participants behave (applying a velocity to the ball), rather than the metric features of the task stimuli.
## Conclusion

@@

::: {.content-visible unless-format="markdown"}
```{dot}
//| label: fig-design-e1
//| fig-cap: "Experiment 1 Design. Constant and Varied participants complete different training conditions."
//| fig-width: 7.0
//| fig-height: 2.5
//| fig-responsive: false
//| echo: false  
///| eval: !expr out_type 
/////| column: screen-inset-right
digraph {
  graph [layout = dot, rankdir = LR]

@@

  Test2 -> Test3
}
```
:::

@@

::: {.content-visible unless-format="markdown"}
```{dot}
//| label: fig-design-e2
//| fig-cap: "Experiment 2 Design. Constant and Varied participants complete different training conditions. The training and testing bands are the reverse of Experiment 1. "
//| fig-width: 8.0
//| fig-height: 2.5
//| fig-responsive: false
//| echo: false 
//| 
/////| column: screen-inset-right
digraph {
  graph [layout = dot, rankdir = LR]
  node [shape = rectangle, style = filled]
  data1 [label = " Varied Training \n100-300\n350-550\n600-800", fillcolor = "#FF0000"]
  data2 [label = " Constant Training \n600-800", fillcolor = "#00A08A"]
  Test3 [label = "    Final Test \n  Novel With Feedback  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
  data1 -> Test1
  data2 -> Test1
  subgraph cluster {
    label = "Test Phase \n(Counterbalanced Order)"
    Test1 [label = "Test  \nNovel Bands  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
    Test2 [label = "  Test \n  Varied Training Bands  \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]
    Test1 -> Test2
  }
  Test2 -> Test3
}
```

@@

::: {.content-visible unless-format="pdf"}
# References

::: {.content-visible unless-format="markdown"}
### Appendix - Project 1

{{< pagebreak >}}

### Appendix - Project 2 - Experiment 1

{{< include Appendix/E1_Appendix.qmd >}}
:::

::: {.content-visible when-format="pdf"}

### 2024-04-06 Sections/project_transition.qmd
render full in pdf and markdown

>>>>

In Project 2, a modified version of the task from Project 1 is used in conjunction with a testing procedure that challenges participants to extrapolate well beyond their training experience. In line with previous research in the function learning literature, participants show evidence of successful extrapolation in our linear task environment. Surprisingly though, the constant training group outperforms the varied training group consistently across numerous variants of the task. Such a pattern is far from unheard of in the vast literature on training variability, and it is therefore remains a worthwhile challenge to evaluate the ability of similarity-based models to account for the observed effects. Additionally, the cognitive process models implemented for project 2 will go beyond the modelling efforts of the previous project in two respects. 1) Extensions that enable the model to produce predictions of participant responses, and 2) fitting and attempting to account for behavior in both training AND testing phases of the experiment.

In project 1, we applied model-based techniques to quantify and control for the similarity between training and testing experience, which in turn enabled us to account for the difference between varied and constant training via an extended version of a similarity based generalization model. In project 2, we will go a step further, implementing a full process model capable of both 1) producing novel responses and 2) modeling behavior in both the learning and testing stages of the experiment. Project 2 also places a greater emphasis on extrapolation performance following training - as varied training has often been purported to be particularly beneficial in such situations. Extrapolation has long been a focus of the literature on function learning (Brehmer 1974; Carroll 1963). Central questions of the function learning literature have included the relative difficulties of learning various functional forms (e.g. linear vs.bilinear vs. quadratic), and the relative effectiveness of rule-based vs. association-based exemplar models vs. various hybrid models (Bott and Heit 2004; DeLosh, McDaniel, and Busemeyer 1997; Jones et al. 2018; Kalish, Lewandowsky, and Kruschke 2004; M. A. Mcdaniel and Busemeyer 2005; M. Mcdaniel et al. 2009). However the issue of training variation has received surprisingly little attention in this area.
